Towards Practical First-Order Model Counting

### 2025-06-12 doc/talk/talk.tex
Start drafting the talk

\documentclass[a4paper,UKenglish,cleveref,autoref,table]{lipics-v2021}
\pdfoutput=1
\bibliographystyle{plainurl}

\title{Towards Practical First-Order Model Counting}

\author{Ananth K. Kidambi}{Indian Institute of Technology Bombay, Mumbai, India}{210051002@iitb.ac.in}{}{}
\author{Guramrit Singh}{Indian Institute of Technology Bombay, Mumbai, India}{guramrit@iitb.ac.in}{}{}
\author{Paulius Dilkas}{University of Toronto, Toronto, Canada \and Vector Institute, Toronto, Canada \and \url{https://dilkas.github.io/}}{paulius.dilkas@utoronto.ca}{https://orcid.org/0000-0001-9185-7840}{}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, Atlanta, GA, USA \and University of Toronto, Toronto, Canada \and \url{https://www.cs.toronto.edu/~meel/}}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{}

\authorrunning{A.\,K. Kidambi, G. Singh, P. Dilkas, and K.\,S. Meel}
\Copyright{Ananth K. Kidambi, Guramrit Singh, Paulius Dilkas, and Kuldeep S. Meel}

\ccsdesc{Theory of computation~Automated reasoning}
\ccsdesc{Theory of computation~Logic and verification}
\ccsdesc{Mathematics of computing~Combinatorics}

\keywords{First-order model counting, knowledge compilation, lifted inference}

%\relatedversiondetails[cite=DBLP:journals/corr/abs-2502-12278]{Full Version}{https://doi.org/10.48550/arXiv.2502.12278}

\supplementdetails[subcategory={Source Code}]{Software}{https://doi.org/10.5281/zenodo.15583934}

\funding{This research was funded in part by the Natural Sciences and
  Engineering Research Council of Canada (NSERC), funding reference number
  RGPIN-2024-05956, and the Digital Research Alliance of Canada
  (\texttt{\href{https://alliancecan.ca/}{alliancecan.ca}}).}

\acknowledgements{The first two authors contributed equally. Part of the
  research was conducted while all authors were at the National University of
  Singapore. We express our gratitude to the anonymous reviewers for their
  valuable comments and suggestions.}

\nolinenumbers
\hideLIPIcs

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{pifont}

\usetikzlibrary{arrows.meta}
\usetikzlibrary{fit}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.misc}

@@

\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\FO}{$\mathsf{FO}$}
\newcommand{\UFO}{$\mathsf{UFO}^{2} + \mathsf{CC}$}

\SetKwProg{Fn}{Function}{:}{}
\SetKwData{Cache}{Cache}

\DeclareMathOperator{\CR}{CR}
\DeclareMathOperator{\DR}{DR}
\DeclareMathOperator{\Reff}{Ref}
\DeclareMathOperator{\Doms}{Doms}

\sisetup{group-separator = {,}}
\sisetup{range-phrase=--}

\newtheorem{assumption}[theorem]{Assumption}

\crefalias{enumi}{type}
\crefname{type}{Type}{Types}
\creflabelformat{type}{#2\textup{#1}#3}
\crefalias{clause}{equation}
\crefname{clause}{clause}{clauses}
\creflabelformat{clause}{#2\textup{(#1)}#3}
\crefalias{formula}{equation}
\crefname{formula}{sentence}{sentences}
\creflabelformat{formula}{#2\textup{(#1)}#3}

\begin{document}

@@

\begin{abstract}
  First-order model counting (FOMC) is the problem of counting the number of
  models of a sentence in first-order logic. Since lifted inference techniques
  rely on reductions to variants of FOMC, the design of scalable methods for
  FOMC has attracted attention from both theoreticians and practitioners over
  the past decade. Recently, a new approach based on first-order knowledge
  compilation was proposed. This approach, called \textsc{Crane}, instead of
  simply providing the final count, generates definitions of (possibly
  recursive) functions that can be evaluated with different arguments to compute
  the model count for any domain size. However, this approach is not fully
  automated, as it requires manual evaluation of the constructed functions. The
  primary contribution of this work is a fully automated compilation algorithm,
  called \Cranetwo{}, which transforms the function definitions into C++ code
  equipped with arbitrary-precision arithmetic. These additions allow the new
  FOMC algorithm to scale to domain sizes over \num{500000} times larger than
  the current state of the art, as demonstrated through experimental results.
\end{abstract}

\section{Introduction}

\emph{First-order model counting} (FOMC) is the task of determining the number
of models for a sentence in first-order logic over a specified domain. The
weighted variant, WFOMC, computes the total weight of these models, linking
logical reasoning with probabilistic
frameworks~\cite{DBLP:conf/ijcai/BroeckTMDR11}. It builds upon earlier efforts
in weighted model counting for propositional
logic~\cite{DBLP:journals/ai/ChaviraD08} and broader attempts to bridge logic
and
probability~\cite{DBLP:journals/ai/Nilsson86,novak2012mathematical,vsaletic2024graded}.
WFOMC is central to \emph{lifted inference}, which enhances the efficiency of
probabilistic calculations by exploiting
symmetries~\cite{DBLP:conf/ecai/Kersting12}. Lifted inference continues to
advance, with applications extending to constraint satisfaction
problems~\cite{DBLP:journals/jair/TotisDRK23} and probabilistic answer set
programming~\cite{DBLP:journals/ijar/AzzoliniR23}. Moreover, WFOMC has proven
effective at reasoning over probabilistic
databases~\cite{DBLP:journals/debu/GribkoffSB14} and probabilistic logic
programs~\cite{DBLP:journals/ijar/RiguzziBZCL17}. FOMC algorithms have also
facilitated breakthroughs in discovering integer
sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23} and developing recurrence
relations for these sequences~\cite{DBLP:conf/kr/DilkasB23}. Recently, these
algorithms have been extended to perform sampling
tasks~\cite{DBLP:journals/ai/WangPWK24}.

The complexity of FOMC is generally measured by \emph{data complexity}, with a
sentence classified as \emph{liftable} if it admits a polynomial-time solution
relative to the domain size~\cite{DBLP:conf/starai/JaegerB12}. While all
sentences with up to two variables are known to be
liftable~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}, Beame et
al.~\cite{DBLP:conf/pods/BeameBGS15} demonstrated that liftability does not
extend to all sentences, identifying an unliftable sentence with three
variables. Recent work has extended the liftable fragment with additional
axioms~\cite{DBLP:conf/aaai/TothK23,DBLP:journals/ai/BremenK23} and counting
quantifiers~\cite{DBLP:journals/jair/Kuzelka21}, expanding our understanding of
liftability.

FOMC algorithms are diverse, with approaches ranging from \emph{first-order
  knowledge compilation} (FOKC) to cell
enumeration~\cite{DBLP:conf/uai/BremenK21}, local
search~\cite{DBLP:journals/pvldb/NiuRDS11}, and Monte Carlo
sampling~\cite{DBLP:journals/cacm/GogateD16}. FOKC-based algorithms are
particularly prominent, transforming sentences into structured representations
such as circuits or graphs. Even when multiple algorithms can solve the same
instance, FOKC algorithms are known to find polynomial-time solutions where the
polynomial has a lower degree than other
approaches~\cite{DBLP:conf/kr/DilkasB23}. The recently developed ability of a
FOKC algorithm to formulate solutions in terms of recursive
functions~\cite{DBLP:conf/kr/DilkasB23} is also noteworthy since the only other
proposed alternative is to guess recursive
relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. Notable examples of FOKC
algorithms include \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} and its
extension \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}.

The \textsc{Crane} algorithm marked a significant step forward, expanding the
range of sentences handled by FOMC algorithms. However, it had notable
limitations: it required manual evaluation of function definitions to compute
model counts and introduced recursive functions without proper base cases,
making it more difficult to use. To address these shortcomings, we present
\Cranetwo{}, a fully automated FOMC algorithm that overcomes the constraints of
its predecessor. \Cranetwo{} can handle domain sizes over \num{500000} times
larger than those of previous algorithms and simplifies the user experience by
automatically managing base cases and compiling function definitions into
efficient C++ programs.

The paper is structured as follows. \Cref{sec:preliminaries} covers the
necessary preliminaries, including notation, terminology, and background
information about (W)FOMC and FOKC\@. \Cref{sec:main} describes all aspects of
\Cranetwo{}:
\begin{itemize}
  \item its overall structure and interactions with \textsc{Crane},
  \item how we identify a sufficient set of base cases for a recursive function,
  \item how we construct the sentence that describes each base case, and
  \item how we compile function definitions into a C++ program with caching
        mechanisms that ensure efficiency.
\end{itemize}
Then, \cref{sec:experiments} presents our experimental results, demonstrating
\Cranetwo{}'s performance compared to other FOMC algorithms. Finally,
\cref{sec:conclusion} concludes the paper by discussing promising avenues for
future work.

\section{Preliminaries}\label{sec:preliminaries}

We begin this section by describing some notation that we will use throughout
the paper. Then, in \cref{sec:logic,sec:fomc}, we introduce the basic
terminology of first-order logic and formally define (W)FOMC\@.
\Cref{sec:threelogics} provides a brief overview of the different types of
first-order logics commonly used in FOMC literature. \Cref{sec:crane} outlines
the principles of FOKC, particularly in the context of \textsc{Crane}. Finally,
in \cref{sec:algebra}, we introduce the algebraic terminology used to describe
the output of \textsc{Crane}, i.e., functions and equations that define them.

\paragraph*{Notation}
We use $\mathbb{N}_{0}$ to represent the set of non-negative integers. In both
algebra and logic, we write
$S[x_{1} \mapsto y_{1}, x_{2} \mapsto y_{2}, \dots, x_{n} \mapsto y_{n}]$ to
denote the application of a \emph{substitution} to an expression $S$, where all
instances of $x_{i}$ are replaced with $y_{i}$ for all $i = 1, \dots, n$.
Additionally, for any variable $n$ and $a, b \in \mathbb{N}_{0}$, let
$[a \le n \le b] \coloneqq \begin{cases}
  1 & \text{if $a \le n \le b$} \\
  0 & \text{otherwise}
\end{cases}$.

\subsection{First-Order Logic}\label{sec:logic}

This section reviews the basic concepts of first-order logic used in FOKC
algorithms. We focus specifically on the format used internally by
\textsc{ForcLift} and its descendants. See \cref{sec:preprocessing} for a brief
overview of how \Cranetwo{} transforms an arbitrary sentence in first-order
logic into this internal format.

A \emph{term} can be either a variable or a constant. An \emph{atom} can be
either $P(t_{1}, \dots, t_{m})$ (i.e., $P(\mathbf{t})$) for some predicate $P$
and terms $t_{1}, \dots, t_{m}$, or $x=y$ for some terms $x$ and $y$. The
\emph{arity} of a predicate is the number of arguments it takes, i.e., $m$ in
the case of the predicate $P$ mentioned above. We write $P/m$ to denote a
predicate along with its arity. A \emph{literal} can be either an atom (i.e., a
\emph{positive} literal) or its negation (i.e., a \emph{negative} literal). An
atom containing no variables, only constants, is called \emph{ground}. A
\emph{clause} is of the form $\forall x_{1} \in \Delta_{1}\text{.
}\forall x_{2} \in \Delta_{2}\dots\text{ }\forall x_{n} \in \Delta_{n}\text{.
}\phi(x_{1}, x_{2}, \dots, x_{n})$, where $\phi$ is a disjunction of literals
that contain only the variables $x_{1}, \dots, x_{n}$ (and any constants). We
say that a clause is a \emph{(positive) unit clause} if there is only one
literal with a predicate, and it is a positive literal. Finally, a
\emph{sentence} is a conjunction of clauses. Throughout the paper, we will use
set-theoretic notation, interpreting a sentence as a set of clauses and a clause
as a set of literals.

\begin{remark*}
  Conforming to previous work~\cite{DBLP:conf/ijcai/BroeckTMDR11}, the
  definition of a clause includes universal quantifiers for all its variables.
  While it is possible to rewrite the entire sentence with all quantifiers at
  the front, the format we describe has proven convenient for practical use.
\end{remark*}

\subsection{First-Order Model Counting}\label{sec:fomc}

In this section, we will formally define FOMC and its weighted variant. Although
this work focuses on FOMC, computing the FOMC using \Cranetwo{} requires using
WFOMC for sentences with existential quantifiers. For such sentences,
preprocessing (described in \cref{sec:preprocessing}) introduces predicates with
non-unary weights that must be accounted for to compute the correct model count.

\begin{definition}[Structure, model]\label{def:model}
  Let $\phi$ be a sentence. For each predicate $P/n$ in $\phi$, let
  ${(\Delta_{i}^{P})}_{i=1}^{n}$ be a list of the corresponding domains. Let
  $\sigma$ be a map from the domains of $\phi$ to their interpretations as
  finite sets, ensuring the sets are pairwise disjoint and contain the
  corresponding constants from $\phi$. A \emph{structure} of $\phi$ is a set $M$
  of ground literals. It is defined by adding to $M$ either $P(\mathbf{t})$ or
  $\neg P(\mathbf{t})$ for every predicate $P/n$ in $\phi$ and every $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{P})$. (Here, $\prod$ is the
  usual Cartesian product.) A structure is a \emph{model} if it makes $\phi$
  valid.
\end{definition}

\begin{example}[Counting bijections]\label{example:bijections}
  Let us consider the following sentence (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two domains $\Gamma$ and $\Delta$:
  \begin{equation}\label[formula]{eq:bijections}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \end{equation}
  Let $\sigma$ be defined as $\sigma(\Gamma) \coloneqq \{\, 1, 2\,\}$ and
  $\sigma(\Delta) \coloneqq \{\,a, b\,\}$. \Cref{eq:bijections} has two models:
  \[
    \{\, P(1, a), P(2, b), \neg P(1, b), \neg P(2, a) \,\} \qquad \text{and} \qquad \{\, P(1, b), P(2, a), \neg P(1, a), \neg P(2, b) \,\}.
  \]
\end{example}

\begin{definition}[WFOMC instance]\label{def:instance}
  A \emph{WFOMC instance} comprises:
  \begin{itemize}
    \item a sentence $\phi$,
    \item two (rational) \emph{weights} $w^{+}(P)$ and $w^{-}(P)$ assigned to
          each predicate $P$ in $\phi$, and
    \item $\sigma$ as described in \cref{def:model}.
  \end{itemize}
  Unless specified otherwise, we assume all weights are equal to 1.
\end{definition}

\begin{definition}[WFOMC~\cite{DBLP:conf/ijcai/BroeckTMDR11}]
  Given a WFOMC instance $(\phi, w^{+}, w^{-}, \sigma)$ as in
  \cref{def:instance}, the \emph{(symmetric) weighted first-order model count}
  (WFOMC) of $\phi$ is
  \begin{equation}\label{eq:wfomc}
    \sum_{M \models \phi} \prod_{P(\mathbf{t}) \in M} w^{+}(P) \prod_{\neg P(\mathbf{t}) \in M} w^{-}(P),
  \end{equation}
  where the sum is over all models of $\phi$.
\end{definition}

\subsection{The Three Logics of FOMC}\label{sec:threelogics}

\begin{table}[t]
  \centering
  \caption{A comparison of the three logics used in FOMC\@. The
    2\textsuperscript{nd}--5\textsuperscript{th} columns refer to the number of
    sorts, support for constants, the maximum number of variables, and supported
    quantifiers, respectively. The last column lists supported atoms in addition
    to those of the form $P(\mathbf{t})$ for a predicate $P/n$ and an $n$-tuple
    of terms $\mathbf{t}$. Here, $k$ and $m$ are non-negative integers, where
    $m$ depends on the domain size, $P$ is a predicate, and $x$ and $y$ are
    terms.}\label{tbl:logics}
  \begin{tabular}{llclll}
    \toprule
    Logic & Sorts & Constants & Variables & Quantifiers & Add.\ atoms\\
    \midrule
    \rowcolor{gray!10}
    \FO & one or more & \cmark & unlimited & $\forall$, $\exists$ & $x = y$\\
    \Ctwo & \multirow{2}{*}{one} & \multirow{2}{*}{\xmark} & \multirow{2}{*}{two} & $\forall$, $\exists$, $\exists^{= k}$, $\exists^{\le k}$, $\exists^{\ge k}$ & ---\\
    \UFO & & & & $\forall$ & $|P| = m$\\
    \bottomrule
  \end{tabular}
\end{table}

FOMC commonly utilises three types of first-order logic: \FO{}, \Ctwo{}, and
\UFO{}. \Cref{tbl:logics} summarises the key differences among them. \FO{} is
the input format for \textsc{ForcLift} and its extensions \textsc{Crane} and
\Cranetwo{}. \Ctwo{} is often used in the literature on \textsc{FastWFOMC} and
related methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}.
(Note that no algorithm accepts \Ctwo{} as input.) Finally, \UFO{} is the input
format supported by the most recent implementation of
\textsc{FastWFOMC}~\cite{DBLP:conf/kr/TothK24}. All three logics are
function-free, and domains are always assumed to be finite. As usual, we
presuppose the \emph{unique name assumption}, which states that two constants
are equal if and only if they are the same constant~\cite{DBLP:books/aw/RN2020}.

In \FO{}, each term has a designated \emph{sort}, and each predicate $P/n$
corresponds to a sequence of $n$ sorts. Each sort has its corresponding domain.
These assignments to sorts are typically left implicit and follow from the
quantifiers, e.g., $\forall x,y \in \Delta\text{. }P(x, y)$ implies that the
variables $x$ and $y$ have the same sort. On the other hand,
$\forall x \in \Delta\text{. }\forall y \in \Gamma\text{. } P(x, y)$ implies
that $x$ and $y$ have different sorts, and it would be improper to write, for
example, $\forall x \in \Delta\text{. }\forall y \in \Gamma\text{.
} P(x, y) \lor x = y$. \FO{} is also the only logic to support constants,
sentences with more than two variables, and the equality predicate. While we do
not explicitly refer to sorts in the rest of the paper, the many-sorted nature
of \FO{} is paramount to the algorithms presented therein.

\begin{remark*}
  In the case of \textsc{ForcLift} and its extensions, support for a sentence as
  valid input does not imply that the algorithm can compile the sentence into a
  circuit or graph suitable for lifted model counting. However,
  \textsc{ForcLift} compilation always succeeds on any \FO{} sentence without
  constants and with at most two
  variables~\cite{DBLP:conf/nips/Broeck11,DBLP:conf/kr/BroeckMD14}.
\end{remark*}

Compared to \FO{}, \Ctwo{} and \UFO{} lack support for constants, the equality
predicate, multiple domains, and sentences with more than two variables. The
advantage that \Ctwo{} brings over \FO{} is the inclusion of \emph{counting
  quantifiers}. That is, alongside $\forall$ and $\exists$, \Ctwo{} supports
$\exists^{=k}$, $\exists^{\le k}$, and $\exists^{\ge k}$ for any positive
integer $k$. For example, $\exists^{=1} x\text{. }\phi(x)$ means that there
exists \emph{exactly one} $x$ such that $\phi(x)$, and $\exists^{\le 2} x\text{.
}\phi(x)$ means that there exist \emph{at most two} such $x$. \UFO{}, on the
other hand, does not support any existential quantifiers but instead
incorporates \emph{(equality) cardinality constraints}. For example, $|P| = 3$
constrains all models to have \emph{precisely three positive literals with the
  predicate $P$}. Except when stated otherwise, we will write all example
sentences in \FO{}.

\subsection{Crane and First-Order Knowledge Compilation}\label{sec:crane}

As our work builds on \textsc{Crane}, in this section, we will briefly outline
the steps \textsc{Crane} goes through to compile a sentence into a set of
function definitions. We divide the inner workings of the algorithm into two
stages: preprocessing and compilation.

\subsubsection{Preprocessing}\label{sec:preprocessing}

This stage transforms an arbitrary sentence into the format described in
\cref{sec:logic}, primarily by eliminating existential quantifiers via a process
called \emph{Skolemization}~\cite{DBLP:conf/kr/BroeckMD14}. For example, the
first conjunct of \cref{eq:bijections}, i.e.,
\begin{equation}\label[formula]{eq:skolemizationinitial}
  \forall x \in \Gamma\text{. }\exists y \in \Delta\text{. } P(x, y)
\end{equation}
is transformed into
\begin{equation}\label[formula]{eq:skolemization}
  \begin{aligned}
    (\forall x \in \Gamma\text{. } &Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &Z(x) \lor \neg P(x, y))\land{}\\
    (\forall x \in \Gamma\text{. } &S(x) \lor Z(x))\land{}\\
    (\forall x \in \Gamma\text{. } \forall y \in \Delta\text{. } &S(x) \lor \neg P(x, y)),
  \end{aligned}
\end{equation}
where $Z/1$ and $S/1$ are two new predicates with $w^{-}(S) = -1$. The idea
behind Skolemization is that the output sentence has more models, but for every
additional model with a positive weight, there is an equivalent model with a
negative weight. Hence, the WFOMC of
\cref{eq:skolemizationinitial,eq:skolemization} is the same. Moreover, note that
in this example, preprocessing introduces some redundancy, i.e., the second and
third clauses of \cref{eq:skolemization} could be removed. These redundancies
are handled at the beginning of the compilation process described in
\cref{sec:compilation}.

To illustrate how Skolemization preserves the WFOMC, we let $\Gamma = \{\,a\,\}$
and $\Delta = \{\,1, 2\,\}$ and assume that $w^+(P) = w^-(P) = 1$. Here, for
conciseness, we omit negative literals when writing models.
\Cref{eq:skolemizationinitial} has three models: $\{\,P(a, 1)\,\}$,
$\{\,P(a, 2)\,\}$, and $\{\,P(a, 1), P(a, 2)\,\}$, all with weight 1, so the
WFOMC is 3. \Cref{eq:skolemization} has the following models:
\begin{itemize}
  \item $\{\,Z(a)\,\}$,
  \item $\{\,Z(a), S(a)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 1)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 2)\,\}$,
  \item $\{\,Z(a), S(a), P(a, 1), P(a, 2)\,\}$.
\end{itemize}
The first model has a weight of -1 because $w^-(S) = -1$, while all the other
models have a weight of 1. Hence, the WFOMC is $4 - 1 = 3$.

\subsubsection{Compilation}\label{sec:compilation}

After preprocessing, \textsc{Crane} compiles the sentence into the triple
$(\mathcal{E}, \mathcal{F}, \mathcal{D})$, where $\mathcal{E}$ is the set of
equations, and $\mathcal{F}$ and $\mathcal{D}$ are auxiliary maps. $\mathcal{F}$
maps function names to sentences in such a way that the evaluation of the former
corresponds to the FOMC of the latter. $\mathcal{D}$ maps function names and
argument indices to domains. The equations in $\mathcal{E}$ can define an
arbitrary number of functions, one of which (which we will always denote as $f$)
represents the solution to the FOMC problem. Computing the FOMC for specific
domain sizes involves invoking $f$ with those sizes as inputs. $\mathcal{D}$
records this correspondence between function arguments and domains.

\begin{example}\label{example:solution}
  \textsc{Crane} compiles \cref{eq:bijections} for bijection counting into
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a newly introduced domain. (We omit the definition of
  $\mathcal{F}$ as the sentences can become quite verbose.) To compute the
  number of bijections between two sets of cardinality 3, one would evaluate
  $f(3, 3)$; however, the definition of $g$ is incomplete: $g$ is a recursive
  function presented without any base cases. $\mathcal{D}$ encodes that in
  $f(m, n)$, $m$ and $n$ represent $|\Gamma|$ and $|\Delta|$, respectively.
  Similarly, in $g(l, m)$, $l$ represents $|\Delta^{\top}|$, and $m$ represents
  $|\Gamma|$.
\end{example}

Compilation is performed primarily by applying \emph{(compilation) rules} to
sentences. \textsc{Crane} has two modes depending on the selection process for
compilation rules when multiple alternatives are available. The first option is
to use greedy search: there is a list of rules, and the first applicable rule is
the one that gets used, disregarding all the others. The second option is to use
a combination of greedy and \emph{breadth-first search} (BFS). In this approach,
we classify each compilation rule as greedy or non-greedy. Greedy rules are
applied as soon as possible at any stage of the compilation process, while BFS
is executed over all applicable non-greedy rules, identifying the solution that
necessitates the smallest number of non-greedy rules.

\subsection{Algebra}\label{sec:algebra}

In this paper, we use both logical and algebraic constructs. While the rest of
\cref{sec:preliminaries} focused on the former, this section describes the
latter. We write $\expr{}$ for an arbitrary algebraic expression. In the context
of algebra, a \emph{constant} is a non-negative integer. Likewise, a
\emph{variable} can either be a parameter of a function or a variable introduced
through summation, such as $i$ in the expression $\sum_{i=1}^{n} \expr$. A
\emph{function call} is $f(x_{1}, \dots, x_{n})$ (or $f(\mathbf{x})$ for short),
where $f$ is an $n$-ary function, and each $x_{i}$ is an algebraic expression
consisting of variables and constants. A (function) \emph{signature} is a
function call that contains only variables. Given two function calls,
$f(\mathbf{x})$ and $f(\mathbf{y})$, we say that $f(\mathbf{y})$ \emph{matches}
$f(\mathbf{x})$ if $x_{i} = y_{i}$ whenever $x_{i}, y_{i} \in \mathbb{N}_{0}$.
An \emph{equation} is $f(\mathbf{x}) = \expr{}$, where $f(\mathbf{x})$ is a
function call.

\begin{definition}[Base case]\label{def:basecase}
  Let $f(\mathbf{x})$ be a function call where each $x_{i}$ is either a constant
  or a variable. Then the function call $f(\mathbf{y})$ is a \emph{base case} of
  $f(\mathbf{x})$ if $f(\mathbf{y}) = f(\mathbf{x})\sigma$, where $\sigma$ is a
  substitution that replaces one or more \emph{variable} $x_{i}$'s with a
  constant while leaving constants unchanged.
\end{definition}

\begin{example}
  In the equation $f(m, n) = f(m-1, n) + nf(m-1, n-1)$, the only constant is
  $1$, and the variables are $m$ and $n$. The equation contains three function
  calls: one on the left-hand side (LHS) and two on the right-hand side (RHS).
  The function call on the LHS is a signature. Function calls such as $f(4, n)$,
  $f(m, 0)$, and $f(8, 1)$ are all considered base cases of $f(m, n)$ (only some
  of which are useful).
\end{example}

\section{Technical Contributions}\label{sec:main}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \node[anchor=west] at (-1, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compile to C++};

    \node[draw,rounded rectangle,dashed] at (9, -3) (cpp) {C++ code};
    \node[anchor=west] at (-1, -3) (sizes) {Domain sizes};
    \node at (12, -3) (count) {Answer};

    \node[draw,rounded rectangle] at (3, -2) (findbasecases) {\FindBaseCases};
    \node[draw,rounded rectangle,left = 0.1cm of findbasecases] (crane) {\Compile};
    \node[draw,rounded rectangle,right = 0.1cm of findbasecases] (propagate) {\Propagate};
    \node[draw,rounded rectangle,right = 0.1cm of propagate] (simplify) {\Simplify};

    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (0.7, 0.5) {\Cranetwo};

    \draw[-Latex] (formula) -- (compilewithbasecases);
    \draw[-Latex] (compilewithbasecases) -- node[above] {$\mathcal{E}$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);

    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (crane);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (findbasecases);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (propagate);
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,right] {uses} (simplify);
  \end{tikzpicture}
  \caption[]{Using \Cranetwo{} to compute the model count of a sentence $\phi$.
    First, \Cranetwo{} compiles $\phi$ into a set of equations, which form the
    basis for creating a C++ program. Executing the program with different
    command line arguments calculates the model count of $\phi$ for various
    domain sizes. To accomplish this, the \CompileWithBaseCases procedure
    employs \textsc{Crane}, algebraic simplification techniques (denoted as
    \Simplify), and two other auxiliary procedures. }\label{fig:overview}
\end{figure}

\Cref{fig:overview} provides an overview of \Cranetwo{}'s workflow. We will
briefly describe and motivate each procedure before going into more detail in
the corresponding subsection.

\CompileWithBaseCases (see \cref{sec:completing}), the core procedure of
\Cranetwo{}, is responsible for completing the function definitions produced by
\textsc{Crane} with the necessary base cases. To do so, it may recursively call
itself (and \textsc{Crane}) on other sentences. We prove that the number of such
recursive calls is upper bound by the number of domains.

\cref{sec:completing} also describes the \Simplify procedure for algebraic
simplification. It is crucial for simplifying, e.g., a sum of $n$ terms, only
some of which are non-zero. More generally, the equations returned by
\textsc{Crane} often benefit from easily detectable algebraic simplifications
such as $0 \cdot \text{anything} = 0$ and $\text{anything}^{0} = 1$. Note that
the latter includes $0^{0} = 1$, as 1 is the multiplicative identity.

\FindBaseCases (described in \cref{sec:identifying}) inspects a set of equations
to identify a sufficient set of base cases for a given set of equations. We
prove that the returned set of base cases ensures that the evaluation of the
resulting function definitions will never get stuck in an infinite loop.

>>>>

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{tikz}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}

\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{fit}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.misc}

@@

\beamertemplatenavigationsymbolsempty
\usetheme{default}
\usecolortheme{beaver}

\author{Ananth K. Kidambi\inst{1} \and Guramrit Singh\inst{1} \and \textbf{Paulius Dilkas}\inst{2,3} \and Kuldeep S. Meel\inst{4,2}}
\institute{\inst{1} IIT Bombay, India \and \inst{2} University of Toronto, Canada \and \inst{3} Vector Institute, Canada \and \inst{4} Georgia Tech, USA}

\title{Towards Practical First-Order Model Counting}
\date{SAT 2025}

% TODO: 20 min / 14 slides (presumably)
\begin{document}

@@

\begin{frame}{Knowledge Compilation Workflow}
  \centering
  \begin{tikzpicture}
    % Top row
    \node[anchor=west] at (-0.5, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (compilewithbasecases) {\CompileWithBaseCases};
    \node[draw,rounded rectangle] at (8, 0) (compilation) {Compile to C++};

    % Bottom row
    \node[draw,rounded rectangle,dashed] at (8, -3) (cpp) {C++ code};
    \node[anchor=west] at (-1, -3) (sizes) {Domain sizes};
    \node at (8, -4) (count) {Model count};

    % Modules
    \node[draw,rounded rectangle,anchor=north west] at (1.2, -1) (crane) {\Compile};
    \node[draw,rounded rectangle,right=0.1cm of crane] (findbasecases) {\FindBaseCases};
    \node[draw,rounded rectangle,anchor=north west] at (1.2, -1.6) (propagate) {\Propagate};
    \node[draw,rounded rectangle,right=0.1cm of propagate] (simplify) {\Simplify};

    % Bounding box and its name
    \node[draw,fit={(compilewithbasecases) (compilation) (crane) (findbasecases) (propagate)},inner ysep=7pt,yshift=5pt] {};
    \node at (1.5, 0.5) {\Cranetwo};

    % Brace and its arrow
    \node[fit=(crane)(findbasecases)(propagate)(simplify)] (uses) {};
    \draw[decorate,decoration={brace}] (uses.north west) -- (uses.north east) node[midway] (brace) {};
    \draw[-Latex,dashed] (compilewithbasecases) -- node[midway,left] {uses} (brace);

    % All other arrows
    \draw[-Latex] (formula) -- (compilewithbasecases);
    \draw[-Latex] (compilewithbasecases) -- node[above] {$\mathcal{E}$} (compilation);
    \draw[-Latex] (compilation) -- (cpp);
    \draw[-Latex] (sizes) -- (cpp);
    \draw[-Latex] (cpp) -- (count);
  \end{tikzpicture} % TODO: maybe add colours?
\end{frame}
